---
title: "Final Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Science 101

This tutorial will take you through the data science pipeline using an example dataset.

## 1. Data Curation, Parsing, and Management

The first part of the pipeline involves obtaining a dataset and modifying it based on the data you want to study.

For example, suppose we want to study teenage suicide rates in the United States.

Here we load a dataset on suicide rates:

```{r load_data, message=FALSE, warning=FALSE}
library(tidyverse)

suicide_df <- read_csv("suicide_rates.csv")

suicide_df
```

To make the dataframe more manageable, we'll remove some of the columns (aka attributes) that we won't be using in our analysis:

```{r}
suicide_df <- suicide_df %>%
  select(country, year, sex, age, suicides_no, population, `suicides/100k pop`, generation)

suicide_df
```

Also, because we are only interested in teenage suicide rates and suicides that occur in the United States, we'll remove all entities with an age outside the 15-24 age range and a country that is not the United States:

```{r}
suicide_df <- suicide_df %>%
  filter(age == "15-24 years" & country == "United States")

suicide_df
```

Now we can graph the data to find trends!

## 2. Exploratory Data Analysis

Let's make a scatter plot of suicides/100k population over the years (we'll also differentiate the male suicides from the female suicides):

```{r}
suicide_df %>%
  ggplot(mapping=aes(y=`suicides/100k pop`, x=year, color=sex)) +
  geom_point()
```

From the graph, it is obvious that the male suicide rate is significantly higher than the female suicide rate.  It also seems that the overall suicide rate increases until 1993-1994, then decreases until 2002-2003, and then increases again until 2015.

## 3. Hypothesis Testing and Machine Learning

Let's try to fit a model to the data. Fitting models is important because it allows us to make estimates on new data points (in this case, the model would tell us what the expected suicide rates would be in future years).  Here, we see if a linear regression model is an appropriate model:

```{r}
suicide_fit <- lm(`suicides/100k pop`~year, data=suicide_df)

suicide_fit_stats <- suicide_fit %>%
  broom::tidy()

suicide_fit_stats
```

Unfortunately, when we only account for the year, a linear regression model is not the best model for this data because the p-value is significantly greater than 0.05 (so we cannot reject the null hypothesis).

However, let's add in sex as a factor:

```{r}
suicide_fit <- lm(`suicides/100k pop`~year+sex, data=suicide_df)

suicide_fit_stats <- suicide_fit %>%
  broom::tidy()

suicide_fit_stats
```

Here, the p-values are significantly less than 0.05, so we can reject the null hypothesis of no relationship between year and sex, and suicide rates.

This shows that holding everything else constant, the number of suicides per 100k population decreases by about 0.11 every year.  Also, holding everything else constant, the number of suicides per 100k population increases by about 15.14 when the sex changes from female to male.

## 4. Conclusions

There are many different trends between various variables that we can observe in any dataframe.  In this suicide rates data in particular, we saw just from a simple scatter plot that teenage male suicide rates are significantly higher than teenage female suicide rates in the United States.